\unnumberedChapter{Summary}

We now have a core library which can handle the basics of just about any node and we have a main controller board that will form the heart of many projects to come for the layout control system. However, while LCS boards will all run the same core library, there are still differences how for example the controller pins are assigned on the boards. It is very likely that new software versions and perhaps different processors and boards layouts will require that the core library is adapted to each processor / board combination. This is especially true for monolithic board designs, where there is a great freedom which pins to use for what. At this point, many projects found on the web will now start a symphony of "ifdefs" in their coding. The conditional compiles often reach up to the highest layers of the firmware programming. This \textbf{controller dependent layer} presented before addresses this issue with offering a mapping structure that will translate between the symbolic names and the actual HW counterpart.

Just before we move on, a good question is what can be done with this board ?. It seems to be just a fancy wrapping around the controller chip. Well, first of all, it is a complete LCS node, capable of processing CAN bus messages. It also features the library runtime storage and exports a variety of pins to the extension connector. During the library development of the concepts and the runtime library this board served as the development platform. When designing a new node type, it is a great starting point to take care of the main controller part, allowing to concentrate fully on the node additional capabilities. We will pick up on this in later chapters when extensions are presented.   