\unnumberedChapter{Introduction}

// intro, motivation ....



LCS hardware modules typically consist of a controller portion and an extension portion. Welcome to the extension part of hardware module design. To recap, controller boards features an extension connector that contains signal lines to the extension board. This connector provides among other lines an I2C bus, which is the communication bus to elements on the extension boards.

Each hardware module almost certainly has additional requirements. Depending on the module type there will be a great variety of extensions. A good example is the sensor and actor hardware for managing the respective turnouts, signals, and so on. A solution to these family of modules is to use the I2C communication bus and a few more controller lines between controller main board and extensions. Rather than exporting as many as possible pins of the controller via an extension connector, the extension connector presented is hopefully a reasonable compromise when it comes to numbers of pins and capabilities. For complex extension boards with many controller interaction points, a monolithic approach, i.e. controller and extension functionality on one board, would perhaps be the better choice. Such boards could still export the basic functions such as I2C via the extension connector to less complex extension boards for providing additional capabilities. Before we discuss any particular extension board, this chapter will present the hardware and firmware foundation for addressing an extension board.



// ??? what to put here ? overview on the modules ? 